Global adjoint tomography -- Model GLAD-M25\footnote{This paper is dedicated to Dimitri Komatitsch, computational geophysicist extraordinaire, who passed away unexpectedly shortly before its submission.}

### 2019-02-14 main.tex
Updates from Overleaf

%%---------------------------------------------
%
%   Summary
%
%%---------------------------------------------
\begin{summary}
Building on our first-generation global adjoint tomography model,
GLAD-M15 \citep{bozdaug2016global}, we present our second-generation transversely isotropic global model,
named GLAD-M25, which is the result of ten quasi-Newton tomographic iterations with an
earthquake database consisting of 1,480 events in the magnitude range $5.5\le M_w \le 7.2$,
an almost six-fold increase over the first-generation model.
We calculated fully 3D synthetic seismograms with a shortest period of 17~s based on a GPU-accelerated spectral-element wave
propagation solver which accommodates effects due to 3-D anelastic crust \& mantle structure, topography \& bathymetry, the ocean load, ellipticity, rotation, and self-gravitation.
We used an adjoint-state method to calculate Fr\'echet derivatives in 3D anelastic Earth models
facilitated by a parsimonious storage algorithm.
The simulations were performed on the Cray XK7 ``Titan'' at the Oak Ridge Leadership Computing Facility.
We quantitatively evaluated GLAD-M25 by assessing misfit reductions and traveltime \& amplitude anomaly histograms in twelve measurement
categories.
We performed similar assessments for a held-out data set consisting of 360 earthquakes,
with results comparable to the actual inversion.
We compare GLAD-M25 with numerous other global and regional models,
and highlight a variety of plumes and subduction zones.
\end{summary}

@@

Before using 1,040 earthquakes in our structural inversion,
we performed CMT inversions using model GLAD-M15,
the results of which are summarized in Fig.~\ref{fig:source_correction}.
To ensure even global coverage,
we used seismographic stations from the  Global Seismic Network (II, IU, IC, US, CU, and GT),
GEOFON~(GE), GEOSCOPE~(G), and several regional networks, such as MedNet~(MN),
the Brazilian Lithospheric Seismic Project~(BL), the Chilean National Seismic Network~(C),
and the Japan Meteorological Agency Seismic Network~(JP) (Fig.~\ref{fig:source_correction}b).
The number of stations used for CMT inversion usually ranges
from~150 to~500.

@@

\begin{figure}[ht!]
    \centering
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/Afar.png}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/Bermuda_Canary.png}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/CapeVerde_Hoggar.png}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/Iceland_Eifel.png}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/Easter_Galapagos.png}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/Marion3_Kerguelen.png}}\\
    \caption{\small{Vertical cross sections for various plumes for models GLAD-M25 ($V_\textrm{P}$; left column), TX2015~\citep[$V_\textrm{S}$;][]{TX2015} (center column), and SEMUCB-WM1~\citep[$V_\textrm{S}$;][]{french2015broad} (right column).
    The map in the top left corner of each row shows the cross section with color-coded red, green, and white dots for geographical reference;
    hotspots are denoted by red triangles.
    The dashed black semicircles in the cross sections denote depths of 410~km, 660~km, and 1000~km.
    (a) Afar; (b) Bermuda (left) and Canary (middle); (c) Cape Verde (middle) and Hoggar (right); (d) Iceland (middle) and Eifel (right); (e) Easter (left) and Galapagos (right); (f) Marion (middle), and Kerguelen (right). }}
    \label{fig:plumes}
\end{figure}

@@

\begin{figure}
    \centering
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Aeg.png}\label{fig:subd_a}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Br.png}\label{fig:subd_b}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Ha.png}\label{fig:subd_c}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Wc.png}\label{fig:subd_e}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Ne.png}\label{fig:subd_d}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Su.png}\label{fig:subd_f}}\\[-1pt]
    \caption{\small{Vertical cross sections of compressional wavespeed perturbations in various subduction zones
    for models GLAD-M25 ($V_\textrm{P}$; left column), GAP-P4~\citep{fukao2013subducted} (middle column) and UU-P07~\citep{van2018atlas} (right column).
    The map in the top left corner of each row shows the cross section with color-coded red, green, and white dots for geographical reference.
    The dashed black semicircles in the cross sections denote depths of 410~km, 660~km, and 1000~km.
    (a) Aegean; (b) South America; (c) Hatteras; (d) Wichita. (e) Nepal; (f) Sunda. \red{(GLAD-M25 is still Vp)}}}
    \label{fig:subd}
\end{figure}

In this section we highlight various subduction systems on Earth.
In Fig.~\ref{fig:subd} we compare GLAD-M25~($V_\textrm{P}$) with two compressional wavespeed models,
namely, GAP-P4~\citep{fukao2013subducted} and UU-P07~\citep{van2018atlas}.
Fig.~\ref{fig:subd}(a) features the Aegean subduction zone. All models show a slab penetrating into the lower mantle beyond a depth of 1,000~km.
Fig.~\ref{fig:subd}(b) shows the South American slab, which, in this cross section, ponds in the shallow lower mantle in all three models, before sinking all the way to the CMB.
Figs.~\ref{fig:subd}(c) and~(d) show cuts across the ancient Farallon slab, which penetrates deeply into the lower mantle,
as spectacularly documented by~\cite[e.g.]{grand1994,hilst1997,grand1997high}.
At depths greater than 1,000~km, 
we see a very strong expression of the slab in model GLAD-M25, which is much weaker in the two P models.
The difference may be attributed to increased data coverage thanks to USArray, as well as the use of finite-frequency Fr\'echet derivatives.
Fig.~\ref{fig:subd}(e) shows subduction beneath the Himalayas.
All three models show two distinct fast anomalies below 660~km,
one above 1,000~km and the other, further to the southwest, below 1,000~km.
Finally, Fig.~\ref{fig:subd}(f) shows subduction beneath the Java portion of the Sunda Arc.
At this location the slab pernetrates into the lower mantle, at which point its spreads out laterally,
in agreement with models GAP-P4 and UU-P07.
\cite{fukao1992} explained this flattening using a model of~\cite{ringwood1988},
suggesting that the subducted slab thickened and buckled to form a dense megalith
above the 660~km discontinuity before sinking into the lower mantle.

@@

Seismology is a science driven by data.
There was a time when seismologists would meticulously inspect single seismograms to glean valuable information from them.
Today, we have easy access to data from thousands of earthquakes recorded by tens of thousands of seismographic stations, and individual inspection of all of these records has become impossible.
Still, it would be a shame not to utilize as much as possible of the information in this massive seismographic database.

>>>>

%%---------------------------------------------
%
%   Summary
%
%%---------------------------------------------
\begin{summary}
Building on our first-generation global adjoint tomography model,
GLAD-M15~\citep{bozdaug2016global}, we present our second-generation transversely isotropic global model,
named GLAD-M25, which is the result of ten quasi-Newton tomographic iterations with an
earthquake database consisting of 1,480 events in the magnitude range $5.5\le M_w \le 7.2$,
an almost six-fold increase over the first-generation model.
We calculated fully 3D synthetic seismograms with a shortest period of 17~s based on a GPU-accelerated spectral-element wave
propagation solver which accommodates effects due to 3-D anelastic crust \& mantle structure, topography \& bathymetry, the ocean load, ellipticity, rotation, and self-gravitation.
We used an adjoint-state method to calculate Fr\'echet derivatives in 3D anelastic Earth models
facilitated by a parsimonious storage algorithm.
The simulations were performed on the Cray XK7 ``Titan'' at the Oak Ridge Leadership Computing Facility.
We quantitatively evaluated GLAD-M25 by assessing misfit reductions and traveltime \& amplitude anomaly histograms in twelve measurement
categories.
We performed similar assessments for a held-out data set consisting of 360 earthquakes,
with results comparable to the actual inversion.
We compare GLAD-M25 with numerous other global and regional models,
and highlight a variety of plumes and subduction zones.
\end{summary}

@@

Before using the 1,040 earthquakes in our structural inversion,
we performed CMT inversions using model GLAD-M15,
the results of which are summarized in Fig.~\ref{fig:source_correction}.
To ensure even global coverage,
we used seismographic stations from the  Global Seismic Network (II, IU, IC, US, CU, and GT),
GEOFON~(GE), GEOSCOPE~(G), and several regional networks, such as MedNet~(MN),
the Brazilian Lithospheric Seismic Project~(BL), the Chilean National Seismic Network~(C),
and the Japan Meteorological Agency Seismic Network~(JP) (Fig.~\ref{fig:source_correction}b).
The number of stations used for CMT inversion usually ranges
from~150 to~500.

@@

\begin{figure}[ht!]
    \centering
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/Afar.png}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/Bermuda_Canary.png}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/CapeVerde_Hoggar.png}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/Iceland_Eifel.png}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/Easter_Galapagos.png}}\\[-1pt]
    \sidesubfloat[]{\includegraphics[width=0.98\textwidth]{figures/plumes/Marion3_Kerguelen.png}}\\
    \caption{\small{Vertical cross sections for various plumes for models GLAD-M25 ($V_\textrm{S}$; left column), TX2015~\citep[$V_\textrm{S}$;][]{TX2015} (center column), and SEMUCB-WM1~\citep[$V_\textrm{S}$;][]{french2015broad} (right column).
    The map in the top left corner of each row shows the cross section with color-coded red, green, and white dots for geographical reference;
    hotspots are denoted by red triangles.
    The dashed black semicircles in the cross sections denote depths of 410~km, 660~km, and 1000~km.
    (a) Afar; (b) Bermuda (left) and Canary (middle); (c) Cape Verde (middle) and Hoggar (right); (d) Iceland (middle) and Eifel (right); (e) Easter (left) and Galapagos (right); (f) Marion (middle), and Kerguelen (right). \red{(GLAD-M25 is still Vp)}}}
    \label{fig:plumes}
\end{figure}

@@

\begin{figure}
    \centering
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Aeg.png}\label{fig:subd_a}}\\[-8pt]
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Br.png}\label{fig:subd_b}}\\[-8pt]
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Ha.png}\label{fig:subd_c}}\\[-8pt]
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Wc.png}\label{fig:subd_e}}\\[-8pt]
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Ne.png}\label{fig:subd_d}}\\[-8pt]
    \sidesubfloat[]{\includegraphics[width=0.88\textwidth]{figures/subductions/Su.png}\label{fig:subd_f}}\\[-8pt]
    \caption{\small{Vertical cross sections of compressional wavespeed perturbations in various subduction zones
    for models GLAD-M25 ($V_\textrm{P}$; left column), GAP-P4~\citep{fukao2013subducted} (middle column) and UU-P07~\citep{van2018atlas} (right column).
    The map in the top left corner of each row shows the cross section with color-coded red, green, and white dots for geographical reference.
    The dashed black semicircles in the cross sections denote depths of 410~km, 660~km, and 1000~km.
    (a) Aegean; (b) South America; (c) Hatteras; (d) Wichita. (e) Nepal; (f) Sunda.}}
    \label{fig:subd}
\end{figure}

In this section we highlight various subduction systems on Earth.
In Fig.~\ref{fig:subd} we compare GLAD-M25~($V_\textrm{P}$) with two compressional wavespeed models,
namely, GAP-P4~\citep{fukao2013subducted} and UU-P07~\citep{van2018atlas}.
Fig.~\ref{fig:subd}(a) features the Aegean subduction zone. All models show a slab penetrating into the lower mantle beyond a depth of 1,000~km.
Fig.~\ref{fig:subd}(b) shows the South American slab, which, in this cross section, ponds in the shallow lower mantle in all three models, before sinking all the way to the CMB.
Figs.~\ref{fig:subd}(c) and~(d) show cuts across the ancient Farallon slab, which penetrates deeply into the lower mantle,
as spectacularly documented by, for instance, \citet{grand1994,hilst1997,grand1997high}.
At depths greater than 1,000~km, 
we see a very strong expression of the slab in model GLAD-M25, which is much weaker in the two P models.
The difference may be attributed to increased data coverage thanks to USArray, as well as the use of finite-frequency Fr\'echet derivatives.
Fig.~\ref{fig:subd}(e) shows subduction beneath the Himalayas.
All three models show two distinct fast anomalies below 660~km,
one above 1,000~km and the other, further to the southwest, below 1,000~km.
Finally, Fig.~\ref{fig:subd}(f) shows subduction beneath the Java portion of the Sunda Arc.
At this location the slab pernetrates into the lower mantle, at which point its spreads out laterally,
in agreement with models GAP-P4 and UU-P07.
\cite{fukao1992} explained this flattening using a model of~\cite{ringwood1988},
suggesting that the subducted slab thickened and buckled to form a dense megalith
above the 660~km discontinuity before sinking into the lower mantle.

@@

Seismology is a science driven by data.
There was a time when seismologists would meticulously inspect single seismograms to glean valuable information from them \red{(Perhaps it is now the case for planetary seismology? )}.
Today, we have easy access to data from thousands of earthquakes recorded by tens of thousands of seismographic stations, and individual inspection of all of these records has become impossible \red{(probably it depends on the seismologist. I remember Jeroen Ritsema telling that it is not a big deal to go through the entire database in a week or so with some not-so-sophisticated plotting techniques. Those interested in data still look at, I think..)}.
Still, it would be a shame not to utilize as much as possible of the information in this massive seismographic database.
